---
title: "Homework 2"
author: "Harrison Wei (hw3206)"
output: github_document
---

```{r setup, message=FALSE}
library(dplyr)
library(tidyr)
library(readr)
library(readxl)
library(janitor)
```

## Problem 1: NYC subway entrances

Each row in the source file describes an entrance or exit, rather than a whole station. I kept the station identifiers and location, the routes, and the entrance, vending, and accessibility fields needed below. The route columns need to be character: some contain train letters, while sparsely populated columns can otherwise be read as numbers.

```{r subway-import}
subway <- read_csv(
  "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
  na = c("", "NA"),
  show_col_types = FALSE
) |>
  clean_names() |>
  select(
    line, station_name, station_latitude, station_longitude,
    route1:route11, entry, vending, entrance_type, ada
  ) |>
  mutate(
    across(route1:route11, as.character),
    entry = case_when(entry == "YES" ~ TRUE, entry == "NO" ~ FALSE),
    ada = as.logical(ada)
  )
```

The cleaned table has `r nrow(subway)` entrance/exit records and `r ncol(subway)` variables. It includes the station's line and name, coordinates, up to 11 routes, whether entry is allowed, vending availability, entrance type, and ADA status. It is not fully tidy yet: the numbered route columns represent repeated values of one variable.

I count a station by its **line and station name together**, as the question specifies. A station is counted as ADA compliant when at least one of its entrance/exit records is marked ADA compliant.

```{r subway-summaries}
station_count <- subway |>
  distinct(line, station_name) |>
  nrow()

ada_station_count <- subway |>
  filter(ada) |>
  distinct(line, station_name) |>
  nrow()

no_vending_entrances <- subway |>
  filter(vending == "NO")

no_vending_entry_share <- mean(no_vending_entrances$entry)
```

There are **`r station_count` distinct stations**, of which **`r ada_station_count`** have an ADA-compliant entrance/exit. Among the `r nrow(no_vending_entrances)` entrances/exits without vending, `r sum(no_vending_entrances$entry)` allow entry, a proportion of **`r round(no_vending_entry_share, 3)`** (`r round(100 * no_vending_entry_share, 1)`%).

Next I put the route slot number and the train name in separate columns. I drop empty route slots, which are not train services.

```{r subway-routes}
subway_routes <- subway |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name",
    values_drop_na = TRUE
  ) |>
  mutate(route_number = as.integer(sub("route", "", route_number))) |>
  filter(route_name != "")

a_stations <- subway_routes |>
  filter(route_name == "A") |>
  distinct(line, station_name)

a_ada_stations <- subway_routes |>
  filter(route_name == "A", ada) |>
  distinct(line, station_name)
```

The A train serves **`r nrow(a_stations)` distinct stations** in these data; **`r nrow(a_ada_stations)`** of those have an ADA-compliant entrance/exit.

## Problem 2: Trash Wheels

I used the **October 2026** workbook specified in the October 6 course announcement. For each sheet, the import range starts at its real header row and excludes note columns. The rows without a dumpster number are totals or other non-dumpster information, so I remove them before combining the sheets. The three sheets are cleaned separately, following the order in the assignment.

```{r mr-trash}
trash_file <- "data/202610_Trash_Wheel_Collection_Data.xlsx"

mr_trash <- read_excel(
  trash_file,
  sheet = "Mr. Trash Wheel",
  range = cell_limits(c(2, 1), c(NA, 14))
) |>
  clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    dumpster = as.integer(dumpster),
    year = as.integer(year),
    date = as.Date(date),
    sports_balls = as.integer(round(sports_balls)),
    wheel = "Mr. Trash Wheel"
  )
```

```{r professor-trash}
professor_trash <- read_excel(
  trash_file,
  sheet = "Professor Trash Wheel",
  range = cell_limits(c(2, 1), c(NA, 13)),
  na = c("", "dive")
) |>
  clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    dumpster = as.integer(dumpster),
    year = as.integer(year),
    date = as.Date(date),
    wheel = "Professor Trash Wheel"
  )
```

One Professor Trash Wheel plastic-bottle cell contains the text `dive` in the source workbook. It cannot be treated as a count, so I imported it as missing rather than inventing a value.

```{r gwynnda-trash}
gwynnda_trash <- read_excel(
  trash_file,
  sheet = "Gwynnda the Good Wheel of the W",
  range = cell_limits(c(2, 1), c(NA, 12))
) |>
  clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    dumpster = as.integer(dumpster),
    year = as.integer(year),
    date = as.Date(date),
    wheel = "Gwynnda"
  )

trash_wheels <- bind_rows(mr_trash, professor_trash, gwynnda_trash) |>
  select(wheel, everything())

professor_weight <- trash_wheels |>
  filter(wheel == "Professor Trash Wheel") |>
  summarize(total_tons = sum(weight_tons)) |>
  pull(total_tons)

gwynnda_june_butts <- trash_wheels |>
  filter(wheel == "Gwynnda", year == 2022, month == "June") |>
  summarize(total_butts = sum(cigarette_butts)) |>
  pull(total_butts)
```

The combined data have **`r nrow(trash_wheels)` dumpster records**: `r nrow(mr_trash)` for Mr. Trash Wheel, `r nrow(professor_trash)` for Professor Trash Wheel, and `r nrow(gwynnda_trash)` for Gwynnda. Each record identifies a wheel and a dumpster, with a collection date, weight, volume, and counts of litter types such as plastic bottles and cigarette butts. Some litter types were not recorded for every wheel; these remain missing rather than being set to zero. Professor Trash Wheel collected **`r sprintf("%.2f", professor_weight)` tons** in the available data. Gwynnda collected **`r format(gwynnda_june_butts, big.mark = ",", scientific = FALSE)` cigarette butts** in June 2022.

## Problem 3: MCI and amyloid

Both CSV files begin with a descriptive line above the actual header, so I skip that line. The baseline file uses `.` for an unobserved MCI onset age. Four records show MCI onset at or before their baseline age; those participants do not meet the stated inclusion criterion and are excluded.

```{r mci-baseline}
baseline_all <- read_csv(
  "data/MCI_baseline.csv",
  skip = 1,
  na = c("", "NA", "."),
  show_col_types = FALSE
) |>
  clean_names() |>
  rename(
    baseline_age = current_age,
    mci_onset_age = age_at_onset
  )

baseline <- baseline_all |>
  filter(is.na(mci_onset_age) | mci_onset_age > baseline_age) |>
  mutate(
    sex = factor(sex, levels = c(0, 1), labels = c("Female", "Male")),
    apoe4 = factor(
      apoe4, levels = c(0, 1), labels = c("Non-carrier", "Carrier")
    )
  )

women_carrier_share <- baseline |>
  filter(sex == "Female") |>
  summarize(share = mean(apoe4 == "Carrier")) |>
  pull(share)
```

The raw baseline file contains `r nrow(baseline_all)` people; **`r nrow(baseline)` meet the baseline MCI-free criterion**. Their mean baseline age is **`r round(mean(baseline$baseline_age), 2)` years**. During follow-up, **`r sum(!is.na(baseline$mci_onset_age))`** developed MCI. Among women in the eligible cohort, **`r round(100 * women_carrier_share, 1)`%** are APOE4 carriers. Sex and carrier status are labeled factors, not numeric codes.

The amyloid file has one column for each scheduled time point (baseline and years 2, 4, 6, and 8). I pivot those columns into a time variable and a ratio variable, keeping missing measurements visible. That gives one row per participant per scheduled time point.

```{r amyloid-import}
amyloid_wide <- read_csv(
  "data/mci_amyloid.csv",
  skip = 1,
  na = c("", "NA"),
  show_col_types = FALSE
) |>
  clean_names() |>
  rename(id = study_id)

amyloid <- amyloid_wide |>
  pivot_longer(
    baseline:time_8,
    names_to = "visit",
    values_to = "amyloid_ratio"
  ) |>
  mutate(
    years_since_baseline = case_when(
      visit == "baseline" ~ 0L,
      visit == "time_2" ~ 2L,
      visit == "time_4" ~ 4L,
      visit == "time_6" ~ 6L,
      visit == "time_8" ~ 8L
    )
  ) |>
  select(id, years_since_baseline, amyloid_ratio)
```

This file contains **`r nrow(amyloid_wide)` participants** and `r nrow(amyloid)` scheduled participant-visits. Of those visits, `r sum(!is.na(amyloid$amyloid_ratio))` have a recorded amyloid ratio; the rest are missing measurements.

```{r mci-join}
baseline_only <- anti_join(baseline, distinct(amyloid, id), by = "id")
amyloid_only <- anti_join(distinct(amyloid, id), baseline, by = "id")

mci_amyloid <- inner_join(baseline, amyloid, by = "id") |>
  arrange(id, years_since_baseline)

write_csv(mci_amyloid, "data/mci_amyloid_clean.csv")
```

After the baseline eligibility filter, **`r nrow(baseline_only)` people** appear only in the baseline file (IDs `r paste(baseline_only$id, collapse = ", ")`), and **`r nrow(amyloid_only)`** appear only in the amyloid file (IDs `r paste(amyloid_only$id, collapse = ", ")`). Four of the latter have baseline records but were excluded for MCI at baseline; the other 12 have no baseline record. The inner join retains **`r n_distinct(mci_amyloid$id)` participants** and **`r nrow(mci_amyloid)` scheduled participant-visits**. Each row has baseline demographics, follow-up MCI onset age if applicable, the scheduled visit year, and its amyloid ratio; `r sum(is.na(mci_amyloid$amyloid_ratio))` ratios are missing. The exported tidy table is `data/mci_amyloid_clean.csv`.
